Statistical paradoxes in coronavirus case-counts: selection bias, measurement error, and the COVID-19 pandemic

### 2021-07-29 write-up/covid-writeup.tex
Update covid-writeup.tex

\subsection*{Decision-making: data versus information}

Having read through the above technical discussions, one may come to the incorrect conclusion that because we are concerned about implicit biases in observational data, we must wait for better data sources in order to act~\citep{Ioannidis2020}.  This conclusion confuses data and information.  Lack of high quality data means we should be skeptical regarding conclusions drawn from these data sources alone.  Being skeptical of observed data does not imply governments and communities should wait for high quality data to act; this is especially true in the current high risk scenario.  Instead, we should aim to have high quality information.  Yet in the current crisis we have just that.  Indeed, epidemiologists and public health officials have a strong understanding of the basic facts of pandemics and disease spread.  Social distancing, contact tracing and mass testing are all important.  Individuals and communities with low quality data and only basic information will hopefully make decisions that respect inherent risks to their well being as well as the high degree of uncertainty in the observed data. Of course, there are trade-offs and long-term stay-at-home orders will have negative economic consequences.  A full discussion of these topics is beyond the scope of the current article.

@@

Governments implicitly argue that increased testing capacity will perform the above task.  Without complete compliance, however, our understanding of future outbreaks may be plagued by self-selection bias, compounded by changing sensitivity and specificity rates. Random testing removes these effect modifiers, giving governments more information to fight the disease.  This is not to say random testing should replace current testing practices; only that the random testing should supplement and will provide valuable additional information even with moderate sample sizes.

\subsection*{High quality information on prevalence.}

Governments are pushing for increased testing capacity and robust contact tracing.  Contact tracing is incredibly useful for identifying carriers early and preventing spread of the disease.  We would argue that, after the current wave, understanding active infections is also key.  We know that the probability of an outbreak is a function of current prevalence and network connectivity.  Epidemic critical thresholds have been derived for many infectious disease models~\citep{Pastor2001,Newman2002,Parshani2010}.  Knowing prevalence will help governments determine long-term community-level risk and allow for more targeted interventions -- shutting down only certain locations when necessary.

@@

\subsection*{Model-based solutions.}

One can extend the SIR model to account for selection bias and measurement error. Without strong assumptions on the selection mechanism, however, the estimates are often not identifiable.  When an issue ``cannot be resolved nonparametrically then it is usually dangerous to resolve it parametrically''\citep{CoxHink74}. Absent random sampling, the best route forward for all data analyses is careful associated sensitivity analyses and humility in data-driven conclusions.

>>>>

@@

Governments and policy makers often implicitly argue that increased testing capacity will alleviate selection bias.  Without complete compliance, however, our understanding of future outbreaks may be plagued by self-selection bias, compounded by changing sensitivity and specificity rates. Random testing removes these effect modifiers, giving governments more information to fight the disease.  This paper emphasizes that data quanitity is secondary to data quality and methodological considerations to account for selection bias.

% \subsection*{High quality information on prevalence.}

% Governments are pushing for increased testing capacity and robust contact tracing.  Contact tracing is incredibly useful for identifying carriers early and preventing spread of the disease.  We would argue that, after the current wave, understanding active infections is also key.  We know that the probability of an outbreak is a function of current prevalence and network connectivity.  Epidemic critical thresholds have been derived for many infectious disease models~\citep{Pastor2001,Newman2002,Parshani2010}.  Knowing prevalence will help governments determine long-term community-level risk and allow for more targeted interventions -- shutting down only certain locations when necessary.

@@

\subsection*{Purely model-based solutions.}

A common argument is that the SIR model to account for selection bias and measurement error; therefore, there is no need for auxiliary random samples.  Without strong assumptions on the selection mechanism, however, the estimates are often not identifiable.  When an issue ``cannot be resolved nonparametrically then it is usually dangerous to resolve it parametrically''\citep{CoxHink74}. Absent some type of random sampling, the best route forward for all data analyses is careful associated sensitivity analyses and humility in data-driven conclusions.
